Reconstructing the Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-10-31 doc/paper-lcr.tex
checked though up to testing on mocks

Transforming an observed photometometric galaxy catalogue into a 
catalogue of halo masses, positions, and redshifts, will enable us to
attempt a reconstruction of the convergence induced by every halo near a
line of sight. This is the subject of Section \ref{subsec:halos}. We
also need to account for the mass in the Universe that is not associated
with the galaxies we can see -- this is the subject of Section
\ref{subsec:voids}.

@@

\subsection{Halos}
\label{subsec:halos}

@@

This PDF contains a hidden assumption of an uninformative prior PDF for
$kappah$ -- in the case of infinitely poorly measured stellar masses and
redshifts, $\pr(\kappah|\data)$ will have very long tails that we do not
belive to exist. The resolution of this is to apply an additional prior
on $\kappah$, given by the global underlying $\kappah$ distribution from
the simulation, $\pr(\kappah)$. In the limiting  case of very poor
photometric data, $\pr(\kappah|\data)$ then defaults to this
distribution. \todo{All}{Discuss this. It ties us to the simulation --
but is not unreasonable. Is the argument about poor data valid? Tom?}

@@

\subsection{Accounting for voids, filaments and dark structures}
\label{subsec:voids}

Whilst halos contribute a positive convergence to $\kappa$, voids contribute
a negative convergence. When a raybundle passes through a  region of space
that is less dense than $\rho_{\rm cr}(z)$ the rays are de-focused\footnote{In
principle, the convergence caused by halos, is also caused by the overdensity,
but we neglect this effect since halos are typically much more dense than the
intergalactic medium}. Whilst the  full raytraced solution takes into account
the effect of voids, our reconstruction cannot -- the halo catalogue does not
include voids. In principle the absence of a halo could be taken to infer a
void, but the under-density of such a void is hard to infer. Neglecting voids
would lead to a heavily biased estimate of $\kappa$.

The \proceedure outlined in Section \ref{subsec:halos} allows us to estimate 
the convergence due to halos given observations of mass, redshift and position
$\pr(\kappah|\mathcal{D})$, but what we are interested in is the total convergence
along the line of sight $\pr(\kappa|\mathcal{D})$. We can obtain this by
considering the expression:
\begin{equation}
\Pr(\kappa|\mathcal{D}) = \int \dee\kappah 
   \Pr(\kappa|\kappah) \Pr(\kappah|\mathcal{D})
\label{eq:kappaconv}   
\end{equation}
The first term in the integrand relates the summed convergence due to model
halos, $\kappah$, to the true convergence, $\kappa$. In the Millenium
Simulation catalogs, we can compute $\kappah$ for each
selected line of sight and combine multiple lines of sight to calculate $\Pr(\kappa|\kappah)$.

Given an uncertain stellar mass to halo mass conversion, the
$\pr(\kappah|\{\Mstar,z\})$ can shift  relative to
$\pr(\kappah|\{\Mhalo,z\})$; this is because of the highly asymetric
conversion from stellar mass through halo mass into $\kappah$. If this shift
is ignored the resulting $\Pr(\kappa)$ can become highly biased. To calibrate
this effect we reconstruct each of our calibration sightlines, in the same way
as the observed sightlines to generate $\pr(\kappa,\kappah|\{\Mstar,z\})$, for
each calibration sightline we use the median of $\pr(\kappah|\{\Mstar,z\})$ to
form a de-biased $\Pr(\kappa|\kappah)$.

For any given line of sight we can then infer
$\Pr(\kappah|\mathcal{D})$ as described above, and then
multiply it by $\Pr(\kappa|\kappah)$; we do the marginalisation over $\kappah$
by drawing samples from the two-dimensional grid and keeping only the
$\kappa$ values, to form our final result, $\Pr(\kappa|\mathcal{D})$.

@@

\section{Calibrating the halo model}
\label{sec:knownMh+z}

In order to test the reconstruction \proceedure outlined in Section
\ref{sec:model}, we now apply it to mock galaxy catalogs generated in
regions on the sky around lines of sight with known
$\kappa$ in the Millenium Simulation.

@@

\subsection{How does $\kappah$ relate to $\kappa$?}

Given the halo mass catalogue from the Millenium Simulation, 
and convergence maps produced by ray-tracing through the MS's n-body particles, 
it is possible to test the validity of the prescription outlined
in Section \ref{sec:model}. In this section we test reconstruction prescription given perfect
knowledge of the halos' virial masses and redshifts; this provides us with a measure of the intrinsic uncertainty induced by treating matter as truncated NFW halos. 
We test the model by comparing $\kappah$ against
the ray-traced $\kappa$. If the reconstruction process is to be
succesfull the two quantities should be closely related, although
we expect $\kappah$ to be offset from $\kappa$ since the former 
does not include the divergence due to voids. Figure \ref{fig:isitbiased} shows that indeed $\kappah$ is a good estimator, regardless of the individual value of $\kappah$. At fixed $\kappah$ we find that the scatter in $\kapparay$ grows with $\kappah$; our reconstruction is better at reproducing underdense lines of sight than overdense lines of sight.

\begin{figure}
\includegraphics[width=\columnwidth]{figs/cornerplot.eps}
\caption[Biased?]{$\kappah$ verses $\kappa$ for 100000 reconstructed lines of sight. $\kappah$ traces $\kappa$, but with a non-zero offset which is due to the negative convergence from voids. At fixed $\kappah$ the scatter in $\kappa$ grows with $\kappah$.}
\label{fig:isitbiased}
\end{figure}

%\subsubsection{How does $\gammah$ relate to $\gamma$?}

@@

\subsection{How succesful is the calibration at converting $\kappah$ into $\kappa$?}

\begin{figure}
\includegraphics[width=\columnwidth]{figs/globaldist.eps}
\caption[magcut]{Comparison of the global P($\kappa$) from $10^{5}$ lines of sight to $10^{5}$ random samples drawn each from P($\kappax|\{M_{200},z\}$) for a reconstructed lines of sight. Red is the global P($\kappa$) and black is a histogram of the samples. Given perfect knowledge of halo mass and redshift the reconstruction appears to be correct globally.}
\label{fig:globaldist}
\end{figure}

\begin{figure*}
\includegraphics[width=\textwidth]{figs/where_is_the_kappa.eps}
\caption[magcut]{Which objects dominate the external convergence for a line of sight? From left to right the three figures show the cumulative contribution to convergence from indvidual halos as a function of distance from the line of sight, magnitude and redshift. The solid line is the median cumulative contribution, whilst dashed and dot-dashed show the 68 and 95 percent confidence intervals repectively. The majority of convergence comes from halos within the central 30 arcseconds. The convergence is dominated by halos with $i$ magnitudes between 20 and 24, halos brighter than 20 are rare, whilst fainter halos (at $z<1.4$) must be low mass, so can only contribute to the convergence if they are very close to the line of sight. Halos at all redshifts contribute to the convergence but the largest contribution comes from halos with $z \sim z_{\rm source}/2$.}
\label{fig:where}
\end{figure*}

Since voids are not included in the halo model approximation $\kappah$
is a biased estimator of $\kappa$, but this can be calibrated for useing
the \proceedure of Section \ref{subsec:voids}. Taking the reconstructed
$\kappah$ for $1\times 10^{5}$ lines of sight we form the joint
distribution, P$(\kappa,\kappah)$ and marginalize over P$(\kappah)$ to
give P($\kappa$). We define the bias as the difference between the
expectation value of $\kappa$ and the known true value of $\kapparay$
for each Line of sight and we define the reconstruction width,
$\sigma_{\kappa}$,as half the width of the interval containing the
central $68\%$ of the probability P($\kappax$). From $1\times 10^{4}$
reconstructed lines of sight we find the the bias to have mean of
$-9.5\times 10^{-6}$ and a mean reconstruction width of 0.0132 which is
roughly 1.8 times smaller than the width of the global P$(\kappax)$. We
find that our $\kappa$ distributions are consistent with the global
$\kappa$ distribution; Figure \ref{fig:globaldist} shows that sampling
from $\pr(\kappa)$ for 100000 lines of sight is very similar to the
global $\pr(\kappa)$. Given perfect knowledge of halo mass and redshift,
the calibration \proceedure provides an unbiased estimate of $\kappa$
that is $\sim$2 times as precise than using the global $\pr(\kappa)$.

@@

Before investigating the uncertainties caused by imperfect knowledge of
halo mass and redshift, we investigate the uncertainties induced by a
magnitude limited reconstruction and a field of view limited
reconstruction. The halos in our catalogue are populated by galaxies and
given magnitudes according to the semi-analytic model of
\citet{DeLucia+Blaizot2007} and by applying magnitude cuts to our
catalogue, we can investigate the scatter caused by unobserved halos.
The majority of the convergence comes from halos with an $i$ magnitude
between 18 and 24. \Fref{fig:where} shows the cumulative
contribution to $\kappah$ as a function of projected distance from the
line of sight, magnitude and redshift. Unsurprisingly we find that most
of the convergence comes from halos whose galaxies are close to the line
of sight. As a function of magnitude, we find that $\kappah$ is
dominated by objects with magnitudes between $i=18$ and $i=24$. Objects
brighter than $i=18$ are either too rare or too close to the observer to
make a significant contribution to the convergence. Objects fainter than
$i=24$ are too small to be important, unless they are extremely close to
the line of sight, but we find that including including halos fainter
than $i=24$ does not improve our reconstruction. Figure \ref{fig:magcut}
shows the scatter on $\kappah-\kappa$ (where $\kappah$ is shifted so
that it's mean is zero) as a function of reconstruction magnitude limit.
Since these objects must be extremely close to the line of sight to make
a non-zero $\kappah$ contribution it is likely that neglecting stellar
mass and using a spherical NFW prescription are too naive to adequately
reconstruct $\kappax$; thus we do not expect a deeper survey to decrease
the size of the uncertainty in mapping $\kappah$ onto $\kappa$. We find
that halos at all redshifts contribute to the convergence, but halos at
redshifts of $z \sim z_{\rm source}/2$ contribute the most.

@@

For a small fraction of lines of sight, $\pr(\kappax)$ remains very
broad even after applying our reconstruction; these lines of sight are
typically the most overdense lines of sight in the universe. For
time-delay cosmography the most observationally expensive task is the
lightcurve monitoring, whilst making photometric observations of a
4$\times$4 arcminute patch of sky survey down to 24th magnitude is a
relatively cheap. We have shown that with a single epoch observation of
the region around a strong lens it is possible to infer $\pr(\kappax)$.
If a line of sight has a broad $\pr(\kappax)$ it can be rejected {\it
before} the investment of longterm lightcurve monitoring. We find that
selecting only the lines of sight with the most tightly constrained
$\pr(\kappax)$ does not induce a bias.

>>>>

Transforming an observed photometometric galaxy catalogue into a 
catalogue of halo masses, positions, and redshifts, will enable us to
attempt a reconstruction of the convergence induced by every halo near a
line of sight. This is the subject of Section \ref{sec:model:halos}. We
also need to account for the mass in the Universe that is not associated
with the galaxies we can see -- this is the subject of Section
\ref{sec:model:voids}.

@@

\subsection{Halos}
\label{sec:model:halos}

@@

This PDF contains a hidden assumption of an uninformative prior PDF for
$\kappah$ -- in the case of infinitely poorly measured stellar masses and
redshifts, $\pr(\kappah|\data)$ will have very long tails corresponding to
very overdense sightlines that we do not believe exist. The resolution of this
is to divide out the effective prior PDF that was applied during the halo mass
estimation process (which is broad, and hance close to uniform), and apply an
additional prior on $\kappah$, given by the global underlying $\kappah$
distribution from the simulation. In the limiting  case of very poor
photometric data, $\pr(\kappah|\data)$ then defaults to this distribution.
\todo{All}{Discuss this. It ties us to the simulation, but only weakly. Is the
argument about poor data valid? Tom?}

@@

\subsection{Accounting for voids, filaments and other dark structures}
\label{sec:model:voids}

While halos contribute a positive convergence to $\kappa$, voids contribute
a negative convergence. When a ray bundle passes through a  region of space
that is less dense than $\rho_{\rm mean}(z)$ the rays are de-focused\footnote{In
principle, the convergence caused by halos is also caused by the overdensity,
but we neglect this effect since halos are typically much more dense than the
intergalactic medium}. Whilst the  full ray-traced solution takes into account
the effect of voids, our reconstruction cannot -- the halo catalogue does not
include voids. In principle the absence of a halo could be taken to infer a
void, but the under-density of such a void is hard to infer. Neglecting voids
would lead to a heavily biased estimate of $\kappa$. Moreover, the halos in 
the model only add mass to the background density, and that background density
already accounts for all the mass in the Universe! In order to make voids
negative relative to the mean density, we must subtract a significant smooth
mass component. The problem is that we do not have a good model for the
density structure of that smooth component.

The solution we present here is to calibrate the relationship between halo
convergence $\kappah$ and underlying convergence $\kappa$ (as calculated
during full ray tracing) using the simulations themselves.
The halo modelling \proceedure outlined in \Sref{sec:model:halos} allows 
us to estimate 
the convergence due to halos given observations of 
mass, redshift and position,
$\pr(\kappah|\mathcal{D})$, but what we are interested in is the total convergence
along the line of sight $\pr(\kappa|\mathcal{D})$. We can obtain this by
considering the expression:
\begin{equation}
\Pr(\kappa|\mathcal{D}) = \int \dee\kappah 
   \Pr(\kappa|\kappah) \Pr(\kappah|\mathcal{D})
\label{eq:kappaconv}   
\end{equation}
The first term in the integrand relates the convergence due to model halos,
$\kappah$, to the true convergence, $\kappa$. This conditional distribution
can be constructed from the Millennium Simulation catalogs, by computing
$\kappah$ for each selected line of sight using the true values of the halo
parameters,  and accumulating $\kappah$ and $\kappa$ along a large number of
such lines of
sight to  characterize $\Pr(\kappa|\kappah)$.
For any given line of sight we can then estimate
$\pr(\kappah|\mathcal{D})$ as described above, and then
multiply it by $\Pr(\kappa|\kappah)$ and integrate out the intermediate
nuisance parameter $\kappah$. 
% We can do the marginalisation over $\kappah$
% by drawing samples from the two-dimensional grid and keeping only the
% $\kappa$ values, to form our final result, $\Pr(\kappa|\mathcal{D})$.

However, given an uncertain stellar mass to halo mass conversion, the PDF
$\pr(\kappah|\{\Mstar,z\})$ can become skewed relative to
$\pr(\kappah|\{\Mhalo,z\})$; this is because of the assymetry in the
conversion from stellar mass through halo mass into $\kappah$. A long tail of
high halo masses gives rise to a tendency to overestimate $\kappah$.
If this shift is ignored, the resulting $\Pr(\kappa|\data)$ will be 
biased towards high values of $\kappa$.

We find that we can correct this bias to some extent by artificially 
suppressing the tails of $\pr(\kappah|\{\Mstar,z\})$. Let the median of
this PDF be denoted $\kappah^{\rm med}$. We again take a large number of
calibration sightlines from the \MS catalogs, but this time generate a mock
stellar mass catalog for each one. We then reconstruct each lightcone in the
same way as we would an observed field,
estimate $\pr(\kappah|\{\Mstar,z\})$, and extract its median. Accumulating
these samples and their corresponding true values of $kappa$, we can form the 
conditional distribution $\pr(\kappa|\kappah^{\rm med})$. 
Then, when inferring $\kappa$ from an observed photometric catalog, we would
estimate $\pr(\kappah|\{\Mstar,z\}_{\rm obs})$ 
using the halo model, compute the median value 
$\kappa_{\rm h,obs}^{\rm med}$, and then use the following alternative
version of 
\Eref{eq:kappaconv} to infer $\kappa$: 
\bea
\Pr(\kappa|\mathcal{D}) &=& \int \dee\kappah^{\rm med} 
   \Pr(\kappa|\kappah^{\rm med}) \Pr(\kappah^{\rm med}|\mathcal{D}) \notag \\
                        &=& \Pr(\kappa|\kappa_{\rm h,obs}^{\rm med}).
\label{eq:calkappaconv}   
\eea
Here, we have replaced $\Pr(\kappah^{\rm med}|\mathcal{D})$ with a delta
function centred at the observed median value $\kappa_{\rm h,obs}$.

@@

\section{Results: Perfect Halo Data}
\label{sec:knownMh+z}

We now put the reconstruction \proceedure outlined above into practice, first
inferring $\kappa$ from $\kappah$ in the case of hypothetical galaxy catalogs
with noiseless redshifts and halo masses. The reason for doing this is to
check the validity of the calibration process, and then also to  investigate
the primary sources of the convergence. It will also provide us with a measure
of the intrinsic uncertainty introduced by our assumption of all halos being
spherically-symmetric truncated NFW halos with Neto et al concentration--mass
relation.

@@

\subsection{How is $\kappah$ related to $\kappa$?}

\Fref{fig:jointkh-k} shows the joint distribution of $\kappah$ and $\kappa$
values for 100,000 randomly selected \MS lines of sight, assuming noise-free
halo masses and redshifts. In this case, the PDF $\pr(\kappah|\data)$ is a
delta function: to read off $\pr(\kappa|\data)$ we must take a thin horizontal
slice across this joint distribution. Note that the joint PDF plotted in this
figure is the product of  the conditional distribution that we need for
calibration, and also the prior PDF for $\kappah$ that we need to ensure that
the global distribution of $\kappah$ values is defaulted to in the case of
poor observational data. At fixed $\kappah$ we find in \Fref{fig:jointkh-k}
that the scatter in $\kapparay$ grows with $\kappah$; our reconstruction is
better at reproducing underdense lines of sight than overdense lines of sight.
The effect of ignoring voids and the smooth mass component is evident in this
plot: $\kappah$ is significantly higher than $\kappa$, at any given $\kappa$
value. Multiplying by $\pr(\kappah|\data)$ and integrating over $\kappah$
gives a PDF for $\kappa$ centred at the correct place, by construction.

\begin{figure}
\includegraphics[width=\columnwidth]{figs/cornerplot.eps}
\caption[Biased?]{The joint distribution of 
$\kappah$ and $\kappa$, from 100,000 reconstructed lines of
sight. $\kappah$ traces $\kappa$, but with a positive offset which arises from
$\kappah$ not accounting for any voids, dark structures or smooth mass
component. At fixed $\kappah$ the scatter in $\kappa$ grows with $\kappah$.}
\label{fig:jointkh-k}
\end{figure}

@@

\subsection{How accurate is the $\kappah - \kappa$ calibration?}

\begin{figure}
\includegraphics[width=\columnwidth]{figs/globaldist.eps}
\caption[magcut]{Recovering the global $\pr(\kappa)$, shown in red, 
from $10^{5}$ ray-traced lines of
sight. The black curve shows a histogram of 
sample reconstructed 
$\kappa$ values, each one drawn from an inferred PDF 
$P(\kappa|\{\Mhalo,z\}$)
in one of $10^{5}$ reconstructed lines of sight. 
The reconstructions were performed given
perfect knowledge of halo mass and redshift -- in this case
the reconstruction recovers the correct convergence.}
\label{fig:globaldist}
\end{figure}

\begin{figure*}
\includegraphics[width=\textwidth]{figs/where_is_the_kappa.eps}
\caption[magcut]{Which objects dominate the external convergence for a line of
sight? From left to right, the three figures show the cumulative contribution
to the total  convergence from individual halos ($\kappah$) as a function of
1)  distance from the line of sight, 2) magnitude and 3) redshift.   In each
panel, the solid line is the median cumulative contribution over a large
sample of sightlines, while dashed and dot-dashed show the ranges enclosing 68
and 95 percent of the sightlines repectively.}
\label{fig:where}
\end{figure*}

% Since voids are not included in the halo model approximation, $\kappah$
% is a biased estimator of $\kappa$, but this can be calibrated for using
% the \proceedure of \Sref{sec:model:voids}. Taking the reconstructed
% $\kappah$ for $10^{5}$ lines of sight we form the joint
% distribution, $\pr(\kappa,\kappah)$ and marginalize over $\pr(\kappah)$ to
% give $\pr(\kappa)$. 
%
% PJM: WTF? We just went through all of this in the previous section!
% We need to check that we are doing what we say we are.

In this section, let us define the ``bias'' of the reconstruction of a given
line of sight as the difference between the expectation value of $\kappa$ over
the inferred PDF $\pr(\kappa|\data)$ and the known true value $\kappatrue$.
Let us also define the ``width,'' $\sigma_{\kappa}$, to be half the width of
the interval containing the central $68\%$ of the posterior probability
$\pr(\kappa|\data)$.

In an ensemble of $10^{4}$ reconstructed lines of sight, we find the the mean
bias to be $-9.5\times 10^{-6}$, and the mean width to be 0.0132. This is
approximately 1.8 times smaller than the width of the global $\pr(\kappa)$. 
We find that our inferred $\kappa$ PDFs are consistent with the global
$\kappa$ distribution; Figure \ref{fig:globaldist} shows that sampling from
$\pr(\kappa|\data)$ in 100,000 reconstructed lines of sight produces a
distribution of $\kappa$ values over the sky identical to the global
$\pr(\kappa)$. Given perfect knowledge of halo mass and redshift, the
calibration \proceedure provides an unbiased estimate of $\kappa$ that is
$\sim$2 times as precise as using the global $\pr(\kappa)$.

@@

Before investigating the uncertainties caused by imperfect knowledge of halo
mass and redshift, we investigate the uncertainties induced by a limits on the
magnitude of the observed galaxy sample, and on the field of view. The halos
in our catalogue are populated by galaxies and given magnitudes according to
the semi-analytic model of \citet{DeLucia+Blaizot2007}: by applying magnitude
cuts to our catalogue, we can investigate the amount of scatter caused by
unobserved halos.

\Fref{fig:where} shows the cumulative contribution to $\kappah$ as a function
of projected distance from the line of sight, magnitude and redshift.
We find that most of the convergence comes from halos whose
galaxies are close to the line of sight. 
Over half of the  convergence
typically comes from halos within 30 arcseconds of the line of sight; by 2
arcminutes, this fraction is XX\%. \todo{Tom}{Put a number on this.} 
As a function of magnitude, we find
that $\kappah$ is dominated by objects with magnitudes between $i=18$ and
$i=24$. Objects brighter than $i=18$ are either too rare, or too close to the
observer, to make a significant contribution to the convergence. Objects
fainter than $i=24$ are too small to be important, unless they are extremely
close to the line of sight, but we find that including halos fainter than
$i=24$ does not improve the reconstruction.  \Fref{fig:magcut} shows the
scatter on $\kappah-\kappa$ (where $\kappah$ is shifted so that its mean is
zero) as a function of reconstruction magnitude limit. 
%
% Since these objects
% must be extremely close to the line of sight to make a non-zero $\kappah$
% contribution, it is likely that neglecting stellar mass and using a spherical
% NFW prescription are too naive to adequately reconstruct $\kappax$.
%
% PJM: the sentence above does not make sense!
%
We do
not expect a deeper survey to decrease the size of the uncertainty in mapping
$\kappah$ onto $\kappa$. Halos at all redshifts out to the source redshift
(1.4, here) contribute to the convergence, but the largest contribution comes from
halos with $z \sim z_{\rm source}/2$.

@@

For a small fraction of lines of sight, $\pr(\kappax)$ remains very
broad even after applying our reconstruction; these lines of sight are
typically the most overdense lines of sight in the universe. For
time-delay cosmography the most observationally expensive task is the
lightcurve monitoring, whilst making photometric observations of a
4$\times$4 arcminute patch of sky survey down to 24th magnitude is a
relatively cheap. We have shown that with a single epoch observation of
the region around a strong lens it is possible to infer $\pr(\kappax)$.
If a line of sight has a broad $\pr(\kappax)$ it can be rejected {\it
before} the investment of longterm lightcurve monitoring. We find that
selecting only the lines of sight with the most tightly constrained
$\pr(\kappax)$ does not induce a bias.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\includegraphics[width=\columnwidth]{figs/biasplots3.eps}
\caption{}
\label{fig:no-median-tricks}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
